---
title: 'Replicating Point Shaving: Corruption in NCAA Basketball with NBA Data'
author: "Jordan Hutchings"
date: "22/04/2021"
output: rmarkdown::github_document
---

The purpose of this document is to replicate the study, 
[Point Shaving: Corruption in NCAA Basketball](https://users.nber.org/~jwolfers/papers/PointShaving.pdf)

The article author Professor Justin Wolfers, provides betting line data for the 
1991 - 2004 NBA season, similar to the NCAA Div 1 data set used in the paper being 
replicated. The NBA dataset contains 15,507 games while the NCAA dataset contains 
44,120 games from 1989 - 2005. 

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```

```{r, warning=FALSE, message=FALSE}
library(haven)
library(dplyr)
library(ggplot2)
df <- read_dta('betting_odds.dta')
```

## clean the data 
The dataset, while very clean, requires us to construct the spread between the 
difference in scores and the game spread. 

For some reason, the home team names are abreviated, while the away team 
names are in a more readable format. I've created the dictionary `names.csv` 
which maps the home_team names to the away_team names, and used this to update 
the home names.

We are interested in the spread line `odds1`, the result of the spread 
`result1`, and whether the home or away team is favourited. We can deduce 
the better team from the sign of `odds1`, where a positive number indicates 
the home team is favourited, and a negative number favours the away team. 
We can quickly check this by regressing the score difference on the vegas spread. 
A positive coefficient tells us the better the home team is, the higher the spread.

We will construct a variable for the winning margin less the spread, 
and a dummy variable for games with a 12 point spread or greater. 

```{r, message=FALSE, warning=FALSE}
# quick relabeling home teams
names <- read.csv('../names.csv')
df <- merge(df, names, by = 'home_team')
df$home_team <- df$fullname
df <- select(df, -fullname)

# how do we interpret odds1?
df <- mutate(df, score_diff = home_score - away_score) 
# positive beta => the higher odds1, the more favourited team 1 is.
lm(score_diff ~ odds1, data = df) 


# grab favourites, check if favourites won, and by how much relative to line
df <- mutate(df, 
             fav = ifelse(odds1 > 0, "H", "A"),
             fav_win = ifelse(fav == result1, 1, 0),
             spread_margin = home_score - odds1 - away_score, 
             fav_cover = ifelse((fav == "H" & spread_margin > 0), 1, 
                                ifelse((fav == "A" & spread_margin < 0), 1, 0)), 
             twelve = ifelse(abs(odds1) > 11, "12+ point favourites", "<12 point favourites"))

df <- filter(df, !is.na(odds1))

sd = sd(df$spread_margin)

ggplot(df, aes(spread_margin, fill = twelve)) + geom_density(size = 1) +
  stat_function(fun = function(x) dnorm(x, mean = 0, sd = sd),
                color = "black", linetype = "dotted", size = 1) + 
  xlim(c(-30,30)) + 
  facet_wrap(.~twelve) + 
  guides(fill=guide_legend(title="Vegas Spread")) +
  labs(title = "Probability Distribution: Game Outcomes Relative to the Spread", 
       y = "Density", x = "Winning marin, relative to Vegas spread") + 
  theme_bw()
```

We can see there is a significant dip in the number of teams who did not 
cover the spread for games with a 12 point or greater spread. 

The next piece to examine is why choose a spread of 12 points. Why not 
11 or 13 points? To do this, lets look at the probability of a team covering 
the spread relative to the size of the spread for teams who do and do not 
cover. There is an obvious dropoff after the 12.5 point threshold. Since 
we are using NBA and not NCAA data, it is reasonable to expect the point threshold 
to differ slightly.

```{r, message=FALSE, warning=FALSE}
df %>%
  filter(odds1 > 0) %>%
  group_by(odds1) %>%
  summarise(n = n(), 
            cover_percent = sum(fav_cover)/n, 
            se = sqrt(cover_percent*(1-cover_percent) / n)) %>%
  filter(n > 30) %>%
  ungroup() %>%
  ggplot(aes(x=odds1, y=cover_percent)) + 
      geom_point() + 
      geom_errorbar(aes(ymin = cover_percent - 1.96 * se, 
                        ymax = cover_percent + 1.96 * se)) +
  geom_vline(xintercept = 12, linetype = 'dotted') + 
      labs(title = 'Percentage of games covering the spread by Vegas spread', 
           x = 'Vegas spread (0.5 point increments)', 
           y= 'Percentage of games') + 
      theme_bw()
```

Last, we can look at each team on their own to see how often they are covering 
the spread. The third plot may be the most informative, showing the proportion 
of times each team covered the spread with a less than 12 point spread, and 
with a 12 or greater point spread. 
```{r, message=FALSE, warning=FALSE}
# what is the cover rate for winning teams conditional on the spread?
df %>%
  mutate(favourite = ifelse(fav == "H", home_team, away_team)) %>%
  group_by(favourite) %>%
  summarise(cover_percent = mean(fav_cover),
            n = n(),
            se = sqrt(cover_percent*(1-cover_percent)/n)) %>%
  filter(n > 20) %>%
  ggplot(aes(x=favourite, y=cover_percent)) + 
  geom_point(aes(reorder(favourite, cover_percent, mean), y=cover_percent)) + 
  #geom_errorbar(aes(ymin = cover_percent - se * 1.96, ymax = cover_percent + se * 1.96)) + 
  coord_flip() + 
  labs(title = 'Percentage of games covering the spread by organization', 
           x = 'NBA Team', 
           y= 'Percentage') + 
  theme_bw()

df %>%
  mutate(favourite = ifelse(fav == "H", home_team, away_team)) %>%
  group_by(favourite, twelve) %>%
  summarise(cover_percent = mean(fav_cover),
            n = n(),
            se = sqrt(cover_percent*(1-cover_percent)/n)) %>%
  filter(n > 20) %>%
  ggplot(aes(x=favourite, y=cover_percent)) + 
  geom_point(aes(reorder(favourite, cover_percent, mean), y=cover_percent)) + 
  geom_errorbar(aes(ymin = cover_percent - se * 1.96, ymax = cover_percent + se * 1.96)) + 
  coord_flip() + 
  labs(title = 'Percentage of games covering the spread by organization', 
           x = 'NBA Team', 
           y= 'Percentage') + 
  theme_bw() +
  facet_wrap(.~twelve, scales = 'free')

df %>%
  mutate(favourite = ifelse(fav == "H", home_team, away_team)) %>%
  group_by(favourite, twelve) %>%
  summarise(cover_percent = mean(fav_cover),
            n = n(),
            se = sqrt(cover_percent*(1-cover_percent)/n)) %>%
  filter(n > 20) %>%
  ungroup() %>%
  group_by(favourite) %>%
  mutate(n_team = n()) %>%
  filter(n_team > 1) %>%
  ggplot(aes(x=favourite, y=cover_percent)) + 
  geom_point(aes(reorder(favourite, cover_percent, mean), y=cover_percent, color = twelve), size = 2) + 
  geom_line(alpha = 0.2, size = 1) + 
  coord_flip() + 
  guides(color=guide_legend(title="Vegas Spread")) +
  labs(title = 'Percentage of games covering the spread by organization', 
           x = 'NBA Team', 
           y= 'Percentage') + 
  theme_bw()  
```